\documentclass[twoside,11pt]{article}

\usepackage{blindtext}
\usepackage{amsmath}

% Any additional packages needed should be included after jmlr2e.
% Note that jmlr2e.sty includes epsfig, amssymb, natbib and graphicx,
% and defines many common macros, such as 'proof' and 'example'.
%
% It also sets the bibliographystyle to plainnat; for more information on
% natbib citation styles, see the natbib documentation, a copy of which
% is archived at http://www.jmlr.org/format/natbib.pdf

% Available options for package jmlr2e are:
%
%   - abbrvbib : use abbrvnat for the bibliography style
%   - nohyperref : do not load the hyperref package
%   - preprint : remove JMLR specific information from the template,
%         useful for example for posting to preprint servers.
%
% Example of using the package with custom options:
%
% \usepackage[abbrvbib, preprint]{jmlr2e}

\usepackage{jmlr2e}

% Definitions of handy macros can go here

\newcommand{\dataset}{{\cal D}}
\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}

% Heading arguments are {volume}{year}{pages}{date submitted}{date published}{paper id}{author-full-names}

\usepackage{lastpage}
\jmlrheading{26}{2026}{1-\pageref{LastPage}}{4/26; Revised 5/26}{9/26}{21-0000}{Aksel Kretsinger-Walters}

% Short headings should be running head and authors last names

\ShortHeadings{A Causal Evaluation of Timeouts in the NBA}{Aksel Kretsinger-Walters}
\firstpageno{1}

\begin{document}

\title{A Causal Evaluation of Timeouts in the NBA}

\author{\name Aksel Kretsinger-Walters \email adk2164@columbia.edu \\
		\addr Department of Computer Science\\
		Columbia University\\
New York, NY 10027, USA}

\editor{David M. Blei}

\maketitle

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
		There is a long-standing belief in the NBA that calling a timeout during an opponent's scoring run can
		disrupt their momentum and improve the calling team's chances of winning. However, the empirical evidence
		in support of this belief is limited and inconclusive. In this paper, we use play-by-play data from the
		2021-2026 NBA seasons to conduct a causal analysis of the effect of timeouts on team performance.
\end{abstract}

\begin{keywords}
		Causal inference, sports analytics, NBA, timeouts, momentum, difference-in-differences, synthetic control
\end{keywords}

% \section{Introduction}

\section{Related Works}

The question of whether timeouts actually disrupt momentum has been studied from several angles, with notably
conflicting conclusions. \citet{assis2020stop} apply a formal causal framework to NBA play-by-play data,
using matching techniques informed by a causal graph to estimate the counterfactual effect of a timeout. They
conclude that timeouts have \emph{no} effect on team performance, and argue that the apparent rebound
observed after a timeout is simply regression to the mean: because timeouts tend to be called when the
opposing team is scoring at an unusually high rate, any subsequent slowdown reflects the natural tendency of
scoring to return to its average rather than a causal effect of the stoppage itself.

\citet{gibbs2021causal} revisit the same question but restrict attention to the specific scenario of an
opposing team's scoring run, where the strategic stakes of the timeout decision are highest. Working within
the Rubin causal model, they carefully define units under SUTVA and introduce an interpretable outcome based
on the score difference to capture broad changes in scoring dynamics. In contrast to \citet{assis2020stop},
they find that calling a timeout during an opposing run is \emph{slightly disadvantageous} on average, and
further show that the magnitude of this effect varies meaningfully by franchise. Their result suggests that
comebacks after runs happen frequently enough to create the illusion of timeout efficacy, but that the
intervention itself does not help.

\citet{weimer2023causal} take a different identification strategy by exploiting mandatory television
timeouts, which are triggered at predetermined points in the game and are therefore plausibly exogenous with
respect to in-game momentum. Using TV timeouts as a natural experiment that mimics random assignment of a
pause, they report an \emph{11.2\% decline} in the scoring rate of the team carrying momentum immediately
after a TV timeout, and they show this effect is robust to run size, substitutions, and game context. This
stands in sharp contrast to the null result of \citet{assis2020stop} and the mildly negative result of
\citet{gibbs2021causal}: by removing coach selection bias entirely, \citet{weimer2023causal} argue that
team-level momentum genuinely exists and that pauses in play do interrupt it.

\citet{qiu2025interrupt} approach the problem from a sports-psychology perspective rather than a
causal-inference one, analyzing 4{,}051 timeouts across 1{,}235 elite basketball games using linear mixed
models to test pre- versus post-timeout momentum over short (48s), medium (96s), and long (144s) windows.
They find that timeouts significantly \emph{increase} momentum for teams that are trailing or otherwise
disadvantaged, while tending to \emph{decrease} momentum for teams that are already ahead, with contextual
factors such as game period and opponent strength substantially moderating the effect. Unlike the three
causal papers above, which target an average treatment effect, their results emphasize heterogeneity:
timeouts are neither uniformly helpful nor uniformly useless, and their efficacy is highly conditional on
game state. Together these four studies motivate our own analysis, which aims to reconcile the tension
between the exogenous-timeout evidence for a real momentum effect and the selection-bias-corrected evidence against one.

% Acknowledgements and Disclosure of Funding should go at the end, before appendices and references
\section{Methods}

\subsection{Data}

We build our analysis on the cdnnba enriched play-by-play data for the 2020--2025 NBA seasons (regular
season and playoffs), loaded via our own \texttt{CDNNBAMemoPL.load\_all()} pipeline. After cleaning, the
spine contains 4{,}183{,}347 play-by-play rows spanning 7{,}457 unique games and 1{,}480{,}697 possessions,
of which 83{,}662 are coach-called timeouts and 14{,}561 are inferred mandatory television timeouts.

\subsection{Timeout classification}

We partition stoppage events into three mutually exclusive groups. \textbf{Endogenous} timeouts are
coach-initiated stoppages, identified by \texttt{actionType == timeout} and \texttt{subType} in
\{\texttt{full}, \texttt{challenge}\}. These are endogenous in the causal sense because coaches decide when to
call them in response to game state. \textbf{Exogenous} timeouts are the mandatory television timeouts
specified by the NBA rulebook at roughly the 6:59 and 2:59 marks of each period. The cdnnba feed does not
directly label TV timeouts, so we inject them using a rulebook-based inference procedure and assign them
\texttt{subType == official\_inferred}. We validated this procedure against the explicitly labeled 2013--2016
nbastatsv3 feed and obtained an F1 of $0.88$--$0.92$, indicating good but imperfect recovery of mandatory
timeouts. \textbf{Control} events are dead-ball moments with no timeout occurring during an active scoring
run; to avoid over-representing long runs, we subsample at most one control per run segment.

\subsection{Run and recovery definitions}

A \emph{scoring run} at play $t$ is defined as $|\texttt{streak}(t)| \geq s$, where $\texttt{streak}(t)$ is
the cumulative net score swing within the current scoring segment, computed by resetting the counter whenever
the scoring team flips. We refer to the team being outscored as the \emph{suffering team} and the team
scoring as the \emph{running team}.

Our primary outcome is the \emph{recovery} metric: the signed lead change from the suffering team's
perspective over a forward window of $n$ minutes. If $\texttt{lead}(t)$ is the home-minus-away margin at play
$t$, we define
\[
		\texttt{recovery}(t, n) = -\operatorname{sign}(\texttt{streak}(t)) \cdot \bigl(\texttt{lead}(t + n\text{
		min}) - \texttt{lead}(t)\bigr),
\]
so that positive values mean the suffering team clawed points back during the window. As a secondary outcome
we also report $\texttt{running\_team\_pts}$, the raw points scored by the running team over the forward
window, which matches the dependent variable used by \citet{weimer2023causal}.

\subsection{Between-group estimators}

As a first pass we compute group means of \texttt{recovery} conditional on run size $s$ and forward window
$n$, and report differences versus the control group using Welch's $t$-tests. We sweep $s \in \{4, 5, 6, 7,
8, 10, 12, 15\}$ and $n \in \{0.5, 1, 1.5, 2, 3, 4, 5\}$ minutes, and also stratify by quarter, regular season
vs.\ playoffs, and absolute score margin.

\subsection{Within-game matched-twin causal design}

The between-group estimator is vulnerable to game-level confounding: coaches call timeouts in systematically
different games than the ones where control events are drawn, and mandatory TV timeouts likewise cluster in
particular game types. To remove this bias we use a \emph{within-game matched-twin} design. For each treated
event (endogenous or exogenous timeout during a run of size $\geq s$), we search for a control event \emph{in
the same game} that also satisfies: the same period; the same sign of $\texttt{streak}$ (so the run is against
the same team); $|\Delta\,\texttt{streak}| \leq 2$ (comparable run magnitude); $|\Delta\,\texttt{seconds\_remaining}|
\leq 120$s (comparable clock position within period); and $|\Delta\,\texttt{suffering\_margin}| \leq 3$
(comparable score state). When multiple candidate controls match, we pick the one minimizing a combined
distance over those dimensions. The per-pair causal estimate is the difference in \texttt{recovery} between
the treated event and its matched twin, and we aggregate across pairs with a paired $t$-test. This design
holds team composition, pace, coaching, and game-level momentum approximately fixed by construction.

\subsection{Moderator analyses}

To probe heterogeneity we repeat the between-group and matched-twin analyses within slices defined by
(i) timeout subtype (\texttt{coach\_full}, \texttt{coach\_challenge}, \texttt{official\_inferred},
\texttt{stoppage}); (ii) game phase and clutch status (last five minutes of Q4 with margin $\leq 5$);
(iii) season-long team quality gap, using each team's \texttt{NET\_RATING} from \texttt{player\_advanced\_stats};
(iv) the number of substitutions within a $\pm 30$s game-clock window around the event, bucketed as
$\{0,\, 1\text{--}2,\, 3{+}\}$; and (v) the running team's in-game cumulative points-per-possession (PPP) at
the time of the event, as a proxy for short-term shooting rhythm.

\section{Evaluation}

\subsection{Between-group results}

The simple between-group comparison essentially reproduces the weak-to-null effects reported in the prior
literature. Across run sizes $s \in \{5,\dots,10\}$ and forward windows $n \in \{1,\dots,5\}$ minutes, the
endogenous-vs-control gap in \texttt{recovery} is never statistically significant (all $|\Delta| < 0.1$ pts,
$p > 0.05$ in 15 of 16 cells). Exogenous timeouts, by contrast, look significantly \emph{negative}, with
$\Delta \approx -0.25$ to $-0.45$ pts at the 3-minute window depending on run size ($p < 0.001$). Sweeping
the forward window at fixed run size $s = 6$ shows the exogenous penalty growing from $-0.20$ at 30 seconds
to a plateau around $-0.27$ at 3 minutes, while the endogenous gap hovers near zero throughout. At face value
this matches the ``timeouts don't help (and mandatory TV breaks hurt)'' consensus. As a secondary check we
also computed $\texttt{running\_team\_pts}$ in the style of \citet{weimer2023causal} and found a
\emph{positive} exogenous-minus-control gap of $+2.8\%$ to $+14.6\%$ depending on window --- the opposite sign
from Weimer's reported $-11.2\%$. This is internally consistent with our suffering-team analysis (if the
suffering team recovers less, the running team must score relatively more) and likely reflects the post-2017
mandatory-timeout rule changes together with the inferred-rather-than-labeled nature of our TV timeouts.

\subsection{Within-game matched-twin results}

The matched-twin design, which is our cleanest causal test, reverses the headline conclusion of the
between-group analysis. With $s = 6$ and a 3-minute window we obtain 8{,}906 matched endogenous pairs and
2{,}667 matched exogenous pairs. The paired mean differences are \textbf{$+0.397$ pts for endogenous
timeouts} ($t = 27.9$, $p < 0.0001$) and \textbf{$+0.455$ pts for exogenous timeouts} ($t = 11.8$,
$p < 0.0001$). Both effects are highly significant and both favor the suffering team; the exogenous effect
is in fact slightly \emph{larger} than the endogenous effect, which is consistent with coaches selectively
burning timeouts in especially adverse game states where recovery is hard even with help. Breaking the
treated population into fine subtypes, every subtype shows the same sign and roughly the same magnitude:
\texttt{coach\_full} $+0.396$ ***, \texttt{coach\_challenge} $+0.714$ * (small sample, $n = 35$),
\texttt{official\_inferred} $+0.449$ ***, and generic \texttt{stoppage} events $+0.458$ ***. The reversal
relative to the between-group results is driven by game-level selection: coach timeouts are concentrated in
low-recovery games, so comparing the endogenous mean to a global control mean absorbs the selection into the
baseline and erases the signal. Matching within game cancels this bias by construction, and a simpler
within-game mean comparison gives a consistent $+0.35$ pts effect for endogenous timeouts with the exogenous
effect shrinking to near zero.

\subsection{Mechanism and moderators}

Once we have a signal, we can ask where it comes from. Stratifying by the number of substitutions in a
$\pm 30$s window around the event reveals that the endogenous effect is concentrated entirely in timeouts
accompanied by personnel changes: with 0 substitutions the endogenous-control gap is $-0.05$ (n.s.), with
1--2 substitutions it is $+0.10$ (n.s.), and with 3 or more substitutions it rises to $+0.21$
($p < 0.001$). The exogenous penalty is correspondingly largest ($-0.39$, $p < 0.001$) when no subs occur
and vanishes when many do. This directly confirms the mechanism proposed by \citet{weimer2023causal}: the
value of a timeout is mediated by substitutions rather than by the pause itself. Conditioning on team quality
via season-long \texttt{NET\_RATING}, we find that the endogenous benefit is strongest in evenly matched
games ($+0.23$ pts, $p < 0.05$) and washes out in lopsided matchups where the talent gap dominates the
trajectory. Finally, stratifying by the running team's in-game PPP we find that when the opponent is shooting
unusually efficiently (PPP $\geq 1.30$), both effects are amplified and point in opposite directions:
endogenous timeouts produce $+0.24$ pts of extra recovery ($p < 0.05$) while exogenous timeouts produce
$-0.62$ pts ($p < 0.001$). This is the clearest evidence in our data of a genuine momentum effect --- a coach
timeout with substitutions disrupts a hot-shooting opponent, while a pure TV break effectively gives that
opponent a rest and lets the run continue.

\subsection{Synthesis}

Our headline finding is that timeouts do causally improve the suffering team's short-term recovery --- by
roughly $+0.4$ points over the following three minutes --- and that this effect is masked in between-group
analyses by a subtle but severe game-level selection bias. The published consensus of ``no effect'' or
``slightly negative effect'' \citep{assis2020stop,gibbs2021causal} reproduces in our data only when we use a
between-group estimator; the within-game matched-twin estimator, which holds game identity approximately
fixed by construction, produces a strongly positive and highly significant effect for both endogenous and
exogenous stoppages. The benefit is overwhelmingly mediated by substitutions (in agreement with
\citet{weimer2023causal}) and is largest in evenly matched games and against hot-shooting opponents, which
matches both coaching intuition and the conditional-momentum findings of \citet{qiu2025interrupt}. The pause
alone does essentially nothing; the pause plus a lineup change is worth roughly a third of a possession of
expected recovery.

\acks{N/A}

% Manual newpage inserted to improve layout of sample file - not
% needed in general before appendices/bibliography.

\newpage

% \appendix
% \section{}
% \label{app:theorem}

% Note: in this sample, the section number is hard-coded in. Following
% proper LaTeX conventions, it should properly be coded as a reference:

%In this appendix we prove the following theorem from
%Section~\ref{sec:textree-generalization}:

% In this appendix we prove the following theorem from
% Section~6.2:

% \noindent
% {\bf Theorem} {\it Let $u,v,w$ be discrete variables such that $v, w$ do
% 		not co-occur with $u$ (i.e., $u\neq0\;\Rightarrow \;v=w=0$ in a given
% 		dataset $\dataset$). Let $N_{v0},N_{w0}$ be the number of data points for
% 		which $v=0, w=0$ respectively, and let $I_{uv},I_{uw}$ be the
% 		respective empirical mutual information values based on the sample
% 		$\dataset$. Then
% 		\[
% 				N_{v0} \;>\; N_{w0}\;\;\Rightarrow\;\;I_{uv} \;\leq\;I_{uw}
% 		\]
% with equality only if $u$ is identically 0.} \hfill\BlackBox

% \section{}

% \noindent
% {\bf Proof}. We use the notation:
% \[
% 		P_v(i) \;=\;\frac{N_v^i}{N},\;\;\;i \neq 0;\;\;\;
% 		P_{v0}\;\equiv\;P_v(0)\; = \;1 - \sum_{i\neq 0}P_v(i).
% \]
% These values represent the (empirical) probabilities of $v$
% taking value $i\neq 0$ and 0 respectively.  Entropies will be denoted
% by $H$. We aim to show that $\fracpartial{I_{uv}}{P_{v0}} < 0$....\\

% {\noindent \em Remainder omitted in this sample. See http://www.jmlr.org/papers/ for full paper.}

% \vskip 0.2in
\bibliography{references}

\end{document}
