\documentclass[twoside,11pt]{article}

\usepackage{blindtext}
\usepackage{amsmath}
\usepackage{booktabs}

% Any additional packages needed should be included after jmlr2e.
% Note that jmlr2e.sty includes epsfig, amssymb, natbib and graphicx,
% and defines many common macros, such as 'proof' and 'example'.
%
% It also sets the bibliographystyle to plainnat; for more information on
% natbib citation styles, see the natbib documentation, a copy of which
% is archived at http://www.jmlr.org/format/natbib.pdf

% Available options for package jmlr2e are:
%
%   - abbrvbib : use abbrvnat for the bibliography style
%   - nohyperref : do not load the hyperref package
%   - preprint : remove JMLR specific information from the template,
%         useful for example for posting to preprint servers.
%
% Example of using the package with custom options:
%
% \usepackage[abbrvbib, preprint]{jmlr2e}

\usepackage{jmlr2e}
\hypersetup{colorlinks=true, urlcolor=blue, citecolor=blue, linkcolor=blue}

% Definitions of handy macros can go here

\newcommand{\dataset}{{\cal D}}
\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}

% Heading arguments are {volume}{year}{pages}{date submitted}{date published}{paper id}{author-full-names}

\usepackage{lastpage}
\jmlrheading{26}{2026}{1-\pageref{LastPage}}{4/26; Revised 5/26}{9/26}{21-0000}{Aksel Kretsinger-Walters}

% Short headings should be running head and authors last names

\ShortHeadings{A Causal Evaluation of Timeouts in the NBA}{Aksel Kretsinger-Walters}
\firstpageno{1}

\begin{document}

\title{A Causal Evaluation of Timeouts in the NBA}

\author{\name Aksel Kretsinger-Walters \email adk2164@columbia.edu \\
		\addr Department of Computer Science\\
		Columbia University\\
New York, NY 10027, USA}

\editor{David M. Blei}

\maketitle

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
		There is a long-standing belief in the NBA that calling a timeout during an opponent's scoring run can
		disrupt their momentum and improve the calling team's chances of winning. However, the empirical evidence
		in support of this belief is limited and inconclusive. In this paper, we use play-by-play data from the
		2021-2026 NBA seasons to conduct a causal analysis of the effect of timeouts on team performance.
\end{abstract}

\begin{keywords}
		Causal inference, sports analytics, NBA, timeouts, momentum, difference-in-differences, synthetic control
\end{keywords}

\section{Significance}
% Why is this problem important? To whom?

The definition and nature of streaks, runs, and momentum itself has been a point of interest and debate for
sports fans and analysts for decades. In addition to the papers cited in our related works, there's a large
body of research on the existence and nature of momentum in sports, both on the individual level (e.g. the
``hot-hand'' effect) and the team level.

On an academic level, the question of whether timeouts disrupt momentum is a interesting case study in causal
inference in a complex, dynamic system. It involves issues of selection bias (coaches call timeouts in
response to game state), confounding (many factors influence scoring dynamics), and the challenge of defining
and measuring momentum itself. The conflicting conclusions in the existing literature suggest that this is a
non-trivial question.

The significance of this question extends beyond academic curiosity. If timeouts do have a causal effect on
disrupting momentum, it would validate a common coaching strategy and could influence how teams manage their
timeouts. There's also opportunity for further discovery: how does the relationship change conditioned on
player substitutions, home-away status, quarter, regular vs post season, team age and
maturity, duration and intensity of the run, etc.? There are many more such ablations to consider. Conversely,
if timeouts are found to have no effect or even a negative effect, it could lead to a reevaluation of how
coaches use their timeouts and potentially change the way teams approach in-game strategy.

\section{Related Works}

The question of whether timeouts actually disrupt momentum has been studied from several angles, with notably
conflicting conclusions. \citet{assis2020stop} apply a formal causal framework to NBA play-by-play data,
using matching techniques informed by a causal graph to estimate the counterfactual effect of a timeout. They
conclude that timeouts have \emph{no} effect on team performance, and argue that the apparent rebound
observed after a timeout is simply regression to the mean: because timeouts tend to be called when the
opposing team is scoring at an unusually high rate, any subsequent slowdown reflects the natural tendency of
scoring to return to its average rather than a causal effect of the stoppage itself.

\citet{gibbs2021causal} revisit the same question but restrict attention to the specific scenario of an
opposing team's scoring run, where the strategic stakes of the timeout decision are highest. Working within
the Rubin causal model, they carefully define units under SUTVA and introduce an interpretable outcome based
on the score difference to capture broad changes in scoring dynamics. In contrast to \citet{assis2020stop},
they find that calling a timeout during an opposing run is \emph{slightly disadvantageous} on average, and
further show that the magnitude of this effect varies meaningfully by franchise. Their result suggests that
comebacks after runs happen frequently enough to create the illusion of timeout efficacy, but that the
intervention itself does not help.

\citet{weimer2023causal} take a different identification strategy by exploiting mandatory television
timeouts, which are triggered at predetermined points in the game and are therefore plausibly exogenous with
respect to in-game momentum. Using TV timeouts as a natural experiment that mimics random assignment of a
pause, they report an \emph{11.2\% decline} in the scoring rate of the team carrying momentum immediately
after a TV timeout, and they show this effect is robust to run size, substitutions, and game context. This
stands in sharp contrast to the null result of \citet{assis2020stop} and the mildly negative result of
\citet{gibbs2021causal}: by removing coach selection bias entirely, \citet{weimer2023causal} argue that
team-level momentum genuinely exists and that pauses in play do interrupt it.

\citet{qiu2025interrupt} approach the problem from a sports-psychology perspective rather than a
causal-inference one, analyzing 4{,}051 timeouts across 1{,}235 elite basketball games using linear mixed
models to test pre- versus post-timeout momentum over short (48s), medium (96s), and long (144s) windows.
They find that timeouts significantly \emph{increase} momentum for teams that are trailing or otherwise
disadvantaged, while tending to \emph{decrease} momentum for teams that are already ahead, with contextual
factors such as game period and opponent strength substantially moderating the effect. Unlike the three
causal papers above, which target an average treatment effect, their results emphasize heterogeneity:
timeouts are neither uniformly helpful nor uniformly useless, and their efficacy is highly conditional on
game state. Together these four studies motivate our own analysis, which aims to reconcile the tension
between the exogenous-timeout evidence for a real momentum effect and the selection-bias-corrected evidence against one.

% Acknowledgements and Disclosure of Funding should go at the end, before appendices and references
\section{Methods}

\subsection{Data}

We build our analysis on the cdnnba enriched play-by-play data for the 2020--2025 NBA seasons (regular
season and playoffs), loaded via our own \texttt{CDNNBAMemoPL.load\_all()} pipeline. After cleaning, the
spine contains 4{,}168{,}786 play-by-play rows spanning 7{,}457 unique games and 1{,}480{,}697 possessions,
of which 82{,}197 are coach-logged timeouts (\texttt{actionType == timeout}, \texttt{subType} in
\{\texttt{full}, \texttt{challenge}\}). Reclassifying those rows by causal trigger
(\S\ref{sec:timeout-classification}) yields 13{,}172 league-forced \emph{tv\_mandatory} timeouts, 6{,}805
\emph{coach\_absorb} timeouts that satisfy an imminent mandatory slot, 23{,}747
\emph{coach\_discretionary} pure coach calls, 36{,}918 \emph{mistagged\_discretionary} rows (qualifier-tagged
as mandatory by the feed but failing the rulebook's slot-owner / first-team-TO / proximity gates), and
1{,}555 coach challenges.

\subsection{Timeout classification}\label{sec:timeout-classification}

We partition stoppage events into three mutually exclusive treatment groups according to whether the
\emph{coach} made a strategic choice to take the break. \textbf{Exogenous} events are stoppages that would
have occurred regardless of any coach decision: cdnnba \emph{tv\_mandatory} rows (the league-forced
commercial breaks at roughly the 6:59 and 2:59 marks), \emph{coach\_absorb} rows (a coach call within
		$\sim$80 seconds of a pending mandatory trigger, where the slot was queued to auto-fire anyway and the coach
is only choosing the moment, not the fact of the timeout), and natural game pauses (\texttt{actionType ==
stoppage} -- out-of-bounds, injuries, blood rule, equipment, etc.). \textbf{Endogenous} events are
strategic coach calls -- \emph{coach\_discretionary} (no mandatory tag at all),
\emph{mistagged\_discretionary} (qualifier-tagged ``mandatory'' but failing the rulebook gates below), and
coach challenges -- restricted to those called by the team being scored against (the \emph{suffering team}).
\textbf{Control} events are dead-ball moments with no timeout occurring during an active scoring run; to
avoid over-representing long runs, we subsample at most one control per run segment.

The cdnnba feed directly tags mandatory timeouts in its \texttt{qualifiers} column (string contains
``mandatory''), so unlike prior pipelines we do not need to infer the TV-timeout slot from clock position
alone. A qualifier-tagged row becomes \emph{tv\_mandatory} only when (i) \texttt{seconds\_remaining} $\leq$
the period's slot-$K$ trigger and (ii) the charged \texttt{teamId} matches the rulebook's structural slot
owner -- home for slot 1, the opposing team for slot 2. Rows passing only the clock gate (sr $>$ trigger)
become \emph{coach\_absorb} if the coach was within an 80-second buffer of the trigger \emph{and} the row is
that team's first full TO of the period; otherwise the row is reclassified \emph{mistagged\_discretionary}
(the feed billed it as mandatory but it is structurally a coach call). The 80-second buffer is rulebook-
informed but rules-based; we picked it from the empirical knee in the (sr $-$ trigger) distribution and
sensitivity to it is modest in the range 60--180s. To validate this procedure we ran the analogous detector
on the 2013--2016 nbastatsv3 feed -- which carries explicit \texttt{Official} / \texttt{Official TV} ground
truth -- and obtained \textbf{F1 $= 0.933$} (precision $0.876$, recall $0.999$) against that label.

The endogenous/exogenous distinction is causal rather than mechanical. NBA rules allow substitutions and
strategy huddles during \emph{any} timeout (full, short, or mandatory TV), so the opportunity to alter play
after the break is identical in both cases. What separates the two types is how the trigger is selected:
coaches initiate endogenous timeouts in response to game state (introducing selection bias on adverse
momentum), whereas mandatory TV timeouts and coach\_absorb events fire at fixed clock positions and are
therefore plausibly exogenous conditional on the clock.

\subsection{Run and recovery definitions}

A \emph{scoring run} at play $t$ is defined as $|\texttt{streak}(t)| \geq s$, where $\texttt{streak}(t)$ is
the cumulative net score swing within the current scoring segment, computed by resetting the counter whenever
the scoring team flips. We refer to the team being outscored as the \emph{suffering team} and the team
scoring as the \emph{running team} or alternatively the \emph{streaking team}.

Our primary outcome is the \emph{recovery} metric: the signed lead change from the suffering team's
perspective over a forward window of $n$ minutes. If $\texttt{lead}(t)$ is the home-minus-away margin at play
$t$, we define
\[
		\texttt{recovery}(t, n) = -\operatorname{sign}(\texttt{streak}(t)) \cdot \bigl(\texttt{lead}(t + n\text{
		min}) - \texttt{lead}(t)\bigr),
\]
so that positive values mean the suffering team clawed points back during the window. As a secondary outcome
we also report $\texttt{running\_team\_pts}$, the raw points scored by the running team over the forward
window, which matches the dependent variable used by \citet{weimer2023causal}.

\subsection{Between-group estimators}

As a first pass we compute group means of \texttt{recovery} conditional on run size $s$ and forward window
$n$ (measured in minutes), and report differences versus the control group using Welch's $t$-tests. We sweep
$s \in \{4, 5, 6, 7,
8, 10, 12, 15\}$ and $n \in \{0.5, 1, 1.5, 2, 3, 4, 5\}$ minutes, and also stratify by quarter, regular season
vs.\ playoffs, and absolute score margin.

\subsection{Within-game matched-twin causal design}\label{sec:matched-twin}

\paragraph{Motivation.}
The between-group estimator is vulnerable to game-level confounding: coaches call timeouts in systematically
different games than the ones where control events are drawn, and mandatory TV timeouts cluster at
predictable game-clock positions (around the 6:59 and 2:59 marks of each period) rather than uniformly
across all game contexts. Comparing the timeout pool against a global control mean therefore absorbs game-
and clock-level selection into the baseline. To remove this bias we use a \emph{within-game matched-twin}
design that holds team composition, pace, coaching, and game-level momentum approximately fixed by
construction.

\paragraph{Method.}
We first build a candidate pool by classifying every spine event during a run of $|\texttt{streak}| \geq s$
into one of three groups: \textbf{treated-endogenous} (a coach \texttt{full} or \texttt{challenge} timeout
called by the suffering team), \textbf{treated-exogenous} (an \texttt{official\_inferred} TV timeout or a
generic dead-ball \texttt{stoppage}), or \textbf{control} (any spine event during the same run that is not
		classified as treated, with at most one control sampled per contiguous run segment to avoid over-counting
long runs). For each treated event we then search the control pool \emph{within the same game} for a
candidate that satisfies all of: the same period; the same sign of $\texttt{streak}$ (so the run is against
the same team); $|\Delta\,\texttt{streak}| \leq 2$ (comparable run magnitude);
$|\Delta\,\texttt{seconds\_remaining}| \leq 120$\,s (comparable clock position within the period); and
$|\Delta\,\texttt{suffering\_margin}| \leq 3$, where the suffering margin
\[
		\texttt{suffering\_margin}(t) \;=\; -\operatorname{sign}(\texttt{streak}(t)) \cdot \texttt{lead}(t)
\]
is the score margin from the suffering team's perspective (positive when the suffering team is still ahead
despite the run against it). When multiple candidate controls satisfy the constraints, we select the one
minimising the weighted Manhattan distance
\[
		d(\text{treated},\, \text{control}) \;=\; 2\,|\Delta\,\texttt{streak}|
		\;+\; \frac{|\Delta\,\texttt{seconds\_remaining}|}{30}
		\;+\; 3\,|\Delta\,\texttt{suffering\_margin}|,
\]
which trades a ${\sim}30$-second clock gap for one point of streak magnitude or one point of score margin.
Each treated event yields at most one pair; the same control row may serve as the matched twin for multiple
treated events. The per-pair causal estimate is the difference in $\texttt{recovery}$ between the treated
event and its matched twin, and we aggregate across pairs with a paired $t$-test.

\subsection{Moderator analyses}

To probe heterogeneity we repeat the between-group and matched-twin analyses within slices defined by
(i) timeout subtype (\texttt{tv\_mandatory}, \texttt{coach\_absorb}, \texttt{coach\_discretionary},
\texttt{mistagged\_discretionary}, \texttt{coach\_challenge}, \texttt{stoppage});
(ii) game phase and clutch status (last five minutes of Q4 with margin $\leq 5$);
(iii) season-long team quality gap, using each team's \texttt{NET\_RATING} from \texttt{player\_advanced\_stats};
(iv) the number of substitutions within a $\pm 30$s game-clock window around the event, bucketed as
$\{0,\, 1\text{--}2,\, 3{+}\}$; and (v) the running team's in-game cumulative points-per-possession (PPP) at
the time of the event, as a proxy for short-term shooting rhythm.

\section{Evaluation}

\subsection{Between-group results}

The simple between-group comparison essentially reproduces the weak-to-null effects reported in the prior
literature. Across run sizes $s \in \{5,\dots,10\}$ and forward windows $n \in \{1,\dots,5\}$ minutes, the
endogenous-vs-control gap in \texttt{recovery} is never statistically significant (all $|\Delta| < 0.1$ pts,
$p > 0.05$ in 15 of 16 cells). Exogenous timeouts, by contrast, look significantly \emph{negative}, with
$\Delta \approx -0.25$ to $-0.45$ pts at the 3-minute window depending on run size ($p < 0.001$). Sweeping
the forward window at fixed run size $s = 6$ shows the exogenous penalty growing from $-0.20$ at 30 seconds
to a plateau around $-0.27$ at 3 minutes, while the endogenous gap hovers near zero throughout. At face value
this matches the ``timeouts don't help (and mandatory TV breaks hurt)'' consensus. As a secondary check we
also computed $\texttt{running\_team\_pts}$ in the style of \citet{weimer2023causal} and found a
\emph{positive} exogenous-minus-control gap of $+2.8\%$ to $+14.6\%$ depending on window  --  the opposite sign
from Weimer's reported $-11.2\%$. This is internally consistent with our suffering-team analysis (if the
suffering team recovers less, the running team must score relatively more) and likely reflects the post-2017
mandatory-timeout rule changes together with differences in how we classify mandatory timeouts.

To avoid confusion we restate the result: under a between-group estimator, endogenous coach timeouts look
indistinguishable from control and exogenous breaks look mildly negative. We believe this is \textbf{not} a
causal effect but instead a consequence of the adverse situations in which coaches predominantly call
timeouts -- a game-level selection bias that is absorbed into the global control baseline. The matched-twin
design of \S\ref{sec:matched-twin-results}, which holds game identity (and therefore that selection) fixed
by construction, will reverse this conclusion and produce a strongly positive and \emph{symmetric} causal
estimate for both endogenous and exogenous breaks.

\subsection{Within-game matched-twin results}\label{sec:matched-twin-results}

The matched-twin design of \S\ref{sec:matched-twin} -- our cleanest causal test -- reverses the headline
conclusion of the between-group analysis. With $s = 6$ and a 3-minute window we obtain 7{,}509 matched
endogenous pairs and 3{,}480 matched exogenous pairs (each treated event matched to its closest in-game
		control by the weighted distance $d$; control rows are reused across treated rows when they are the best
match for multiple). The paired mean differences are \textbf{$+0.410$ pts for endogenous timeouts}
($t = 26.6$, $p < 0.0001$) and \textbf{$+0.405$ pts for exogenous timeouts} ($t = 13.2$, $p < 0.0001$).
Both effects are highly significant, both favor the suffering team, and -- notably -- they are
\emph{statistically indistinguishable in magnitude}. Breaking the treated population into the six fine
subtypes of \S\ref{sec:timeout-classification} reveals the same sign and roughly the same magnitude
everywhere: \texttt{tv\_mandatory} $+0.361$ *** ($n = 879$), generic \texttt{stoppage} $+0.458$ ***
($n = 1{,}752$), \texttt{coach\_absorb} $+0.340$ *** ($n = 849$), \texttt{coach\_discretionary} $+0.428$ ***
($n = 2{,}119$), \texttt{mistagged\_discretionary} $+0.400$ *** ($n = 5{,}355$), and \texttt{coach\_challenge}
$+0.714$ * ($n = 35$). The consistency across six structurally distinct event types -- league-forced auto-
fires, natural stoppages, last-second coach absorbs, pure discretionary calls, mistagged-as-mandatory coach
calls, and replay challenges -- is itself evidence that the effect is driven by the pause itself, not by
who chose to take it or how the league billed it.

The reversal relative to the between-group results is driven by game-level selection: coach timeouts
concentrate in games where the suffering team historically claws back less of its deficit during opponent
runs (low-recovery games), so comparing the endogenous mean to a global control mean absorbs that selection
into the baseline and erases the signal. The matched twin lives in the same game, same period, with the
same streak sign and a near-identical clock position and score margin, so any game-level confounder (team
quality, pace, coaching style, overall momentum direction) is differenced out at the pair level and the
estimator survives whatever selection-on-game produced the negative exogenous gap in \S4.1. A coarser
within-game \emph{mean} comparison -- in which each treated event is compared against the average
$\texttt{recovery}$ of all controls in the same game, rather than against a single distance-minimising twin
-- gives a consistent $+0.35$ pts effect for endogenous timeouts but shrinks the exogenous effect to near
zero. The drop suggests that tight clock-position matching is more load-bearing for exogenous events (whose
triggers are themselves clock-locked) than for endogenous ones.

The $+0.714$ pair difference on coach challenges is large in magnitude but rests on $n = 35$ matched pairs
and reaches only $p < 0.05$. Two structural factors limit how much weight we can put on this row. First,
the cdnnba feed only emits \texttt{instantreplay} outcome rows for the 2020 and 2021 seasons; challenges
from 2022 onward have no recorded \texttt{overturned} / \texttt{stands} label, so we cannot examine the
broader population. Second, even when we restrict to those two seasons and try to split the matched
challenges by replay outcome (E15), only 19 of the 35 challenges line up with a replay row within our
60-second window and -- coincidentally or otherwise -- all 19 are \texttt{overturned}, leaving zero failed
challenges in the matched sample. The broader 2020--2021 challenge population (1{,}446 challenges from
\texttt{instantreplay/challenge} rows, of which 688 are overturned and 758 fail) shows a near-even
success/failure split, so the all-overturned matched sample is a sampling artifact rather than a true
asymmetry. We therefore cannot test the natural mechanistic hypothesis -- that successful challenges drive
the entire coach\_challenge effect -- with the present matching protocol; we flag it as future work
(\S\ref{sec:future}) for a redesigned challenge-specific match that loosens the per-period clock constraint.

\subsection{Mechanism and moderators}

Once we have a signal, we can ask where it comes from. All moderator analyses in this section use a
between-group Welch test of treated vs control within each stratum -- the same estimator as \S4.1, not the
paired matched-twin of \S4.2. The magnitudes below are therefore directly comparable to the near-zero
between-group baseline and reflect the residual signal that survives the selection bias after stratifying,
not the underlying causal effect size; we use the coarser estimator here to keep per-stratum sample sizes
large enough for a clean test.

Stratifying by the number of substitutions in a $\pm 30$s window around the event reveals that the
endogenous effect is concentrated entirely in timeouts accompanied by personnel changes: with 0
substitutions the endogenous-control gap is $-0.05$ (n.s.), with 1--2 substitutions it is $+0.10$ (n.s.),
and with 3 or more substitutions it rises to $+0.21$ ($p < 0.001$). The exogenous penalty is
correspondingly largest ($-0.39$, $p < 0.001$) when no subs occur and vanishes when many do. This directly
confirms the mechanism proposed by \citet{weimer2023causal}: the value of a timeout is mediated by
substitutions rather than by the pause itself. Conditioning on team quality
via season-long \texttt{NET\_RATING}, we find that the endogenous benefit is strongest in evenly matched
games ($+0.23$ pts, $p < 0.05$) and washes out in lopsided matchups where the talent gap dominates the
trajectory. Finally, stratifying by the running team's in-game PPP we find that when the opponent is shooting
unusually efficiently (PPP $\geq 1.30$), both effects are amplified and point in opposite directions:
endogenous timeouts produce $+0.24$ pts of extra recovery ($p < 0.05$) while exogenous timeouts produce
$-0.62$ pts ($p < 0.001$). This is the clearest evidence in our data of a genuine momentum effect  --  a coach
timeout with substitutions disrupts a hot-shooting opponent, while a pure TV break effectively gives that
opponent a rest and lets the run continue.

\subsection{Summary}

Our headline finding is that timeouts causally improve the suffering team's short-term recovery by roughly
$+0.4$ points over the following three minutes, and that this effect is masked in between-group analyses by
a subtle but severe game-level selection bias. The within-game matched-twin estimator, which holds game
identity approximately fixed by construction, produces a strongly positive and highly significant effect of
\textbf{$+0.410$ pts} for endogenous (strategic coach) breaks and \textbf{$+0.405$ pts} for exogenous
(league-forced or auto-fire) breaks -- two estimates that are statistically indistinguishable. That
symmetry, together with the consistent $+0.34$ to $+0.46$ pair-difference across six structurally distinct
fine subtypes, is strong evidence that the effect is driven by the pause itself rather than by who chose to
take it. The published consensus of ``no effect'' or ``slightly negative effect''
\citep{assis2020stop,gibbs2021causal} reproduces in our data only under a between-group estimator that
absorbs the game-level selection into the baseline. The benefit is overwhelmingly mediated by substitutions
(in agreement with \citet{weimer2023causal}) and is largest in evenly matched games and against hot-shooting
opponents, which matches both coaching intuition and the conditional-momentum findings of
\citet{qiu2025interrupt}.

\section*{Future Work}\label{sec:future}

Several extensions follow naturally from the present design:
\begin{itemize}
		\item \textbf{Better pairing mechanism.} How should we score the quality of a matched twin? The
				current Manhattan-weighted distance $d$ is a reasonable first cut, but there is a sweet-spot
				to characterise between pair similarity (which tightens the estimate) and matched-pair
				sample size (which determines its precision).
		\item \textbf{Alternative outcome metrics.} Beyond $\texttt{recovery}$, we could measure the
				streaking team's points-per-possession or points-per-minute over the forward window, or roll
				either up into a win-probability impact via a calibrated WP model.
		\item \textbf{Causal graphical model.} Distil the moderator findings into an explicit CGM
				(substitutions $\to$ recovery; running-team PPP $\to$ recovery; \texttt{NET\_RATING} gap
				$\to$ recovery) and use it to formalise the identification arguments and check for missing
				confounders.
		\item \textbf{Streaking-team timeouts.} Re-run the analysis with the \emph{streaking} team as the
				calling team. Do they substitute players at a higher rate than baseline, and what mechanism
				would explain a positive or negative effect on the run they are already riding?
		\item \textbf{Additional break types.} Official reviews, injury stoppages, and ejections all
				introduce small exogenous perturbations to play. Do they show the same recovery pattern as
				TV timeouts? Cross-checking against more natural experiments would strengthen the causal
				claim.
\end{itemize}

\section*{Further experimentation and code changes}

A running log of follow-up items raised during the analysis. Items marked \textbf{[done]} have been
folded into the current draft; the rest remain open.

\begin{itemize}
		\item \textbf{[done]} Coach challenge split by successful / failed outcome (E15). Implemented in
				\texttt{run\_experiments\_v2.experiment\_15} and invoked from the notebook;
				results discussed in \S\ref{sec:matched-twin-results}. Matched-sample size collapses to 7
				challenge pairs (19 of which are overturned, 0 failed), so we cannot test the success/failure
				asymmetry with the present per-period match constraint.
		\item \textbf{[done]} Reconcile coach\_challenge counts between the matched-twin section and E15.
				The 35 challenges in the matched sample drop to 19 with replay outcomes attached because the
				cdnnba feed only emits \texttt{instantreplay} rows for the 2020--2021 seasons; from 2022
				onward replay outcomes are absent. Pre-2020 and post-2021 challenges are present as
				timeout rows but cannot be split by outcome.
		\item \textbf{[done]} Mandatory-timeout classifier. The previous inferred-injection pipeline was
				replaced with a cause-based reclassifier driven by the cdnnba \texttt{qualifiers} field,
				with strict rulebook gates on slot owner (home / away) and clock proximity. F1 against the
				v3 ground-truth label rose from $0.88$--$0.92$ to $0.933$.
		\item \textbf{[done]} Reconcile values between sections 4.2 and 4.3 after the new mandatory
				classification. Numbers in both sections are now produced from the same rerun.
		\item \textbf{[done]} Ensure only timeouts called by the suffering team enter the endogenous filter.
				The \texttt{suffering\_called} mask was already present in
				\texttt{build\_rich\_analysis}; verified after the cause-taxonomy refactor.
		\item \textbf{[done]} Re-run the experiment battery after the classifier and cause-taxonomy
				changes, flag differences, update conclusions. The post-rerun headline shifted from
				``endogenous $\approx 0$, exogenous $\approx -0.27$'' (between-group) to ``endogenous and
				exogenous both $\approx +0.4$'' (matched-twin) -- this is now the report's headline.
		\item \emph{(open)} Use real-world elapsed time (\texttt{timeActual}) as a feature in the TV-timeout
				injection model and retrain/evaluate on nbastatsv3. The current cdnnba pipeline does not need
				it (the \texttt{qualifiers} field carries the structural signal directly), but it might
				strengthen the pre-2017 v3 detector.
		\item \emph{(open)} Analyse real-world elapsed time as a mechanism: do longer breaks correlate
				with larger recovery? \texttt{timeout\_duration\_s} is persisted on the dataset and this is
				a one-cell experiment.
		\item \emph{(open)} Team age as a moderator. Use the average age of players on the court at the
				time of the break, plus the delta in average age before vs after the break (a proxy for
				whether the coach swapped out tired veterans for fresh legs).
		\item \emph{(open)} End-of-quarter trimming: never include events in the last $k$ (say, $k = 2$)
				minutes of a period, and only include forward-window outcomes that fit inside the period
				(e.g.\ with 3.5 minutes left, accept the 3-minute window but exclude the 4-minute one).
				Would tighten the controls but cost sample size.
		\item \emph{(open)} Update NBA data through the current season. Expensive: Ted's data is not easily
				redistributable, so this requires re-scraping the cdnnba feed end-to-end.
		\item \emph{(open)} Investigate simple substitutions without official timeouts (e.g.\ at free
				throws). When else are teams allowed to substitute? A clean substitution-without-pause
				control would isolate the personnel-change mechanism from the pause itself.
		\item \emph{(open)} Re-run with the forward window measured in \emph{possessions} instead of
				minutes, ensuring the window contains an even number of possessions so each team gets the
				same opportunity count (modulo the rare technical-foul edge case).
		\item \emph{(open)} Reconcile with the Weimer finding of a $-11.2\%$ drop in streaking-team scoring
				during exogenous breaks. Our suffering-team estimate already inverts the sign in the
				between-group view; a head-to-head replication that mirrors Weimer's exact definition is
				deferred.
\end{itemize}

\acks{N/A}

\section*{LLM Usage}

\texttt{claude\/opus-4.6} was utilized extensively during the implementation of this research project, and we
wanted to acknowledge its contributions.

The author(s) were responsible for project idea generation, sourcing relevant papers and publications,
research roadmap, code scaffolding, and report skeleton.

\texttt{Opus} assisted in the implementation: data preprocessing and cleaning, feature generation,
significance testing, ablation studies, and report completion. All \texttt{Opus} work was closely reviewed
and cross-examined.

% Manual newpage inserted to improve layout of sample file - not
% needed in general before appendices/bibliography.

\newpage

\bibliography{references}

\section*{Appendix}

Tables below reproduce the post-rerun E9--E14 results from the analysis pipeline driver
(\texttt{run\_experiments\_v2.py}) on the cdnnba dataset, run size $s = 6$, 3-minute forward window, with
the cause-based reclassifier and strict slot-owner gate of \S\ref{sec:timeout-classification}. Significance
codes: $\text{n.s.} = p \geq 0.05$, $* = p < 0.05$, $** = p < 0.01$, $*** = p < 0.001$.

\subsection*{E9: Matched-twin causal estimates}

\begin{tabular}{lrrrrr}
		\toprule
		group & pairs & treated $\mu$ & matched ctrl $\mu$ & pair diff & sig \\
		\midrule
		endogenous & 7{,}509 & $+0.352$ & $-0.058$ & $+0.410$ & *** \\
		exogenous  & 3{,}480 & $+0.198$ & $-0.207$ & $+0.405$ & *** \\
		\bottomrule
\end{tabular}

\vspace{0.3em}

\begin{tabular}{lrrrrr}
		\toprule
		subtype & pairs & treated $\mu$ & matched ctrl $\mu$ & pair diff & sig \\
		\midrule
		tv\_mandatory            & 879   & $+0.259$ & $-0.101$ & $+0.361$ & *** \\
		stoppage                 & 1{,}752 & $+0.059$ & $-0.399$ & $+0.458$ & *** \\
		coach\_absorb            & 849   & $+0.422$ & $+0.081$ & $+0.340$ & *** \\
		coach\_discretionary     & 2{,}119 & $+0.352$ & $-0.076$ & $+0.428$ & *** \\
		mistagged\_discretionary & 5{,}355 & $+0.349$ & $-0.051$ & $+0.400$ & *** \\
		coach\_challenge         & 35    & $+0.800$ & $+0.086$ & $+0.714$ & *   \\
		\bottomrule
\end{tabular}

\subsection*{E10: Between-group recovery by fine subtype}

\begin{tabular}{lrrrr}
		\toprule
		subtype & $n$ & $\mu$ & ctrl $\mu$ & $\Delta$ vs ctrl (sig) \\
		\midrule
		tv\_mandatory            & 1{,}486  & $+0.283$ & $+0.371$ & $-0.087$ n.s. \\
		stoppage                 & 3{,}793  & $+0.125$ & $+0.371$ & $-0.246$ ***  \\
		coach\_absorb            & 1{,}563  & $+0.460$ & $+0.371$ & $+0.089$ n.s. \\
		coach\_discretionary     & 4{,}694  & $+0.432$ & $+0.371$ & $+0.062$ n.s. \\
		mistagged\_discretionary & 11{,}543 & $+0.430$ & $+0.371$ & $+0.059$ n.s. \\
		coach\_challenge         & 70       & $-0.429$ & $+0.371$ & $-0.799$ n.s. \\
		\bottomrule
\end{tabular}

\subsection*{E11: Game phase and clutch conditioning}

\begin{tabular}{lrrrrrr}
		\toprule
		condition & endo $n$ & endo $\mu$ & exo $\mu$ & ctrl $\mu$ & $\Delta$ endo-ctrl & $\Delta$ exo-ctrl \\
		\midrule
		Q1 early (0--360s)    & 1{,}415 & $+0.408$ & $+0.327$ & $+0.470$ & $-0.062$ n.s.\ & $-0.143$ n.s. \\
		Q1 late  (360--720s)  & 1{,}242 & $+0.554$ & $+0.322$ & $+0.386$ & $+0.168$ n.s.\ & $-0.063$ n.s. \\
		Q2 early (720--1080s) & 2{,}451 & $+0.585$ & $+0.303$ & $+0.470$ & $+0.114$ n.s.\ & $-0.167$ n.s. \\
		Q2 late  (1080--1440s)& 1{,}524 & $+0.295$ & $+0.148$ & $+0.283$ & $+0.011$ n.s.\ & $-0.136$ n.s. \\
		Q3 early (1440--1800s)& 2{,}517 & $+0.402$ & $+0.289$ & $+0.308$ & $+0.094$ n.s.\ & $-0.018$ n.s. \\
		Q3 late  (1800--2160s)& 1{,}694 & $+0.315$ & $+0.144$ & $+0.407$ & $-0.091$ n.s.\ & $-0.262$ n.s. \\
		Q4 early (2160--2520s)& 2{,}710 & $+0.435$ & $+0.089$ & $+0.317$ & $+0.119$ n.s.\ & $-0.228$ n.s. \\
		Q4 late  (2520--2880s)& 2{,}584 & $+0.390$ & $+0.100$ & $+0.272$ & $+0.118$ n.s.\ & $-0.172$ n.s. \\
		\midrule
		Clutch (Q4 last 5min, $|m|\leq 5$) & 1{,}006  & $+0.257$ & $-0.253$ & $+0.158$ & $+0.100$ n.s.\ & $-0.411$ n.s. \\
		Non-clutch                          & 15{,}301 & $+0.438$ & $+0.249$ & $+0.378$ & $+0.060$ n.s.\ & $-0.129$ *    \\
		\bottomrule
\end{tabular}

\subsection*{E12: Team quality conditioning (\texttt{NET\_RATING})}

\begin{tabular}{lrrrrrr}
		\toprule
		condition & endo $n$ & endo $\mu$ & exo $\mu$ & ctrl $\mu$ & $\Delta$ endo-ctrl & $\Delta$ exo-ctrl \\
		\midrule
		Sufferer much worse ($\Delta \leq -5$)  & 4{,}113 & $-0.153$ & $-0.134$ & $-0.224$ & $+0.071$ n.s.\ & $+0.090$ n.s. \\
		Sufferer worse ($-5 < \Delta \leq -1$)  & 3{,}830 & $+0.312$ & $+0.013$ & $+0.221$ & $+0.091$ n.s.\ & $-0.208$ n.s. \\
		Evenly matched ($|\Delta| < 1$)         & 2{,}409 & $+0.651$ & $+0.170$ & $+0.341$ & $+0.310$ **    & $-0.170$ n.s. \\
		Sufferer better ($1 \leq \Delta < 5$)   & 3{,}345 & $+0.589$ & $+0.453$ & $+0.556$ & $+0.033$ n.s.\ & $-0.103$ n.s. \\
		Sufferer much better ($\Delta \geq 5$)  & 2{,}610 & $+1.096$ & $+0.924$ & $+0.986$ & $+0.110$ n.s.\ & $-0.062$ n.s. \\
		\bottomrule
\end{tabular}

\subsection*{E13: Substitution-adjusted analysis}

\begin{tabular}{lrrrrrr}
		\toprule
		$\pm 30$s subs & endo $n$ & endo $\mu$ & exo $\mu$ & ctrl $\mu$ & $\Delta$ endo-ctrl & $\Delta$ exo-ctrl \\
		\midrule
		0 subs   & 5{,}202 & $+0.424$ & $+0.274$ & $+0.489$ & $-0.065$ n.s.\ & $-0.215$ *    \\
		1--2 subs & 3{,}886 & $+0.342$ & $+0.160$ & $+0.237$ & $+0.105$ n.s.\ & $-0.077$ n.s. \\
		3+ subs   & 7{,}219 & $+0.475$ & $+0.248$ & $+0.262$ & $+0.213$ ***   & $-0.014$ n.s. \\
		\bottomrule
\end{tabular}

\subsection*{E14: Running-team in-game PPP}

\begin{tabular}{lrrrrrr}
		\toprule
		condition & endo $n$ & endo $\mu$ & exo $\mu$ & ctrl $\mu$ & $\Delta$ endo-ctrl & $\Delta$ exo-ctrl \\
		\midrule
		Running PPP $< 1.00$       & 4{,}754 & $+0.345$ & $+0.284$ & $+0.145$ & $+0.200$ *    & $+0.139$ n.s. \\
		Running PPP $1.00$--$1.15$ & 5{,}775 & $+0.420$ & $+0.238$ & $+0.341$ & $+0.079$ n.s.\ & $-0.103$ n.s. \\
		Running PPP $1.15$--$1.30$ & 4{,}171 & $+0.443$ & $+0.255$ & $+0.420$ & $+0.023$ n.s.\ & $-0.164$ n.s. \\
		Running PPP $\geq 1.30$    & 1{,}607 & $+0.656$ & $+0.104$ & $+0.474$ & $+0.182$ n.s.\ & $-0.370$ *    \\
		\bottomrule
\end{tabular}

\end{document}
